R for Clinical Study Reports and Submission

### 2021-07-30 intro.Rmd
first draft of tlf

In clinical trial development, developer needs to organize source code to deliver SDTM or ADaM datasets and table, listing and figures (TLFs).
For example, a Phase 3 clinical trial typically require hundreds of TLFs. 
A consistent and well-defined folder structure is crucial to manage a clinical trial development project.

We recommended to use R package folder structure to organize all the analysis and reporting related source code and documentation
for a clinical trial development project. 
By using the R package folder structure you can follow the convention to organize your code as other contributors on CRAN.
The consistent approach simplify the communication for developers within and across organizations.

@@

- Part 1, "Delivering TLFs in CSR" provides general information with examples to create table, listing and figures. 
In this part, we assume readers are an individual contributor of a clinical project with some experience in R programming. 
Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/) book and 
[R for Data Science](https://r4ds.had.co.nz/) book.

- Part 2, "Clinical Trial Project" provides general information with examples to manage a clinical trial analysis and reporting 
project. In this part, we assume readers are project lead of a clinical project with experience in R package development. 
Some good reference includes [R packages](https://r-pkgs.org/) and [The tidyverse style guide](https://style.tidyverse.org/).

- Part 3, "eCTD Submission package" provides general information in preparing submission package related to CSR in 
electronic Common Technical Document (eCTD). In this part, we assume readers are project lead of a clinical project 
with experience in R package development as well.

@@

We follow the same philosophy described in [Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) and quote here.

- Anything that can be automated, should be automated. 
- Do as little as possible by hand. Do as much as possible with functions. 
- The goal is to spend your time thinking about what you want to do rather than thinking about the minutiae of package structure.

>>>>

In clinical trial development, developer needs to organize source code to deliver SDTM or ADaM datasets and table, listing and figures (TLFs).
For example, a Phase 3 clinical trial typically require hundreds of TLFs. 
A consistent and well-defined folder structure is crucial to manage a clinical trial analysis and reporting (A&R) project.

We recommend to use R package folder structure to organize all the analysis and reporting related source code and documentation
for a clinical trial A&R project. 
By using the R package folder structure you can follow the convention to organize your code as other contributors on CRAN.
The consistent approach simplify the communication for developers within and across organizations.

@@

- Part 1, "Delivering TLFs in CSR" provides general information with examples to create table, listing and figures. 
In this part, we assume readers are an individual contributor of a clinical project with some experience in R programming. 
We expect readers are familiar with data manipulation in R. Some good reference includes [Hands-On Programming with R](https://rstudio-education.github.io/hopr/), 
[R for Data Science](https://r4ds.had.co.nz/) and [Data Manipulation with R](https://www.amazon.com/Data-Manipulation-R-Use/dp/0387747303).

- Part 2, "Clinical Trial Project" provides general information with examples to manage a clinical trial A&R 
project. In this part, we assume a reader is a project lead with experience in R package development. 
Some good reference includes [R packages](https://r-pkgs.org/) and [The tidyverse style guide](https://style.tidyverse.org/).

- Part 3, "eCTD Submission package" provides general information in preparing 
submission package related to clinical study report (CSR) in 
electronic Common Technical Document (eCTD). 
In this part, we assume a reader is a project lead of a clinical project 
with experience in R package development as well.

@@

We share the same philosophy described in [Section 1.1 of R Packages book](https://r-pkgs.org/intro.html?q=Marwick#intro-phil) and quote here.

- "Anything that can be automated, should be automated." 
- "Do as little as possible by hand. Do as much as possible with functions." 
- "The goal is to spend your time thinking about what you want to do 
rather than thinking about the minutiae of package structure."

### 2021-07-30 tlf.Rmd
update title
first draft of tlf

# (PART) Part Delivering TLFs in CSR {-}

@@

ToDo:

add r2rtf related inforamtion

>>>>

# (PART) Delivering TLFs in CSR {-}

@@

## Background

In clinical trial development, it is a critical step to submit information on medicinal products to regulatory agencies. [Electronic Common Technical Document (eCTD)](https://en.wikipedia.org/wiki/Electronic_common_technical_document) 
has become a worldwide regulatory submission standard format. 
For example, [the United States Food and Drug Administration (US FDA) required new drug applications, biologics license applications must be submitted using eCTD format](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd). Clinical Data Interchange Standards Consortium (CDISC) provided a 
[pilot project
following ICH E3 guidance](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse).

Within eCTD, clinical study reports (CSR) are located at module 5. 
[ICH E3 guidance](https://www.ich.org/page/efficacy-guidelines) provide 
compilation of structure and content of clinical study reports.

A typical CSR contain full details of an individual clinical study 
including clinical and statistical description, presentations, and analyses. 
A large number of tables and figures are incorporated into the main text and appendices 
of a CSR. In CDISC pilot project, an [example CSR](file:///C:/Users/zhanyilo/OneDrive%20-%20Merck%20Sharp%20&%20Dohme,%20Corp/Desktop/cdiscpilot01.pdf) is also provided. If you are interested in more examples of clinical study reports, 
the European Medicines Agency (EMA) publishes clinical data including 
clinical study reports submitted by pharmaceutical companies 
on [EMA clincal data website](https://clinicaldata.ema.europa.eu/web/cdp/home).

Building CSR is a team work with clinicians, medical writers, statisticians and statistical programmers.
Here, We focus on the work and deliverables completed by statisticians and statistical programmers. 
In an organization, A group of statisticians and statistical programmers commonly work together to 
define, develop, validate and deliver tables, listings and figures (TLFs) required for a CSR.
Microsoft Word is widely used to prepare CSR in pharmaceutical industry. 
Therefore, `rtf`, `doc`, `docx` are commonly used format to deliver TLFs.

In this chapter, our focus is to illustrate how to create tables, listings and figures (TLFs) in RTF format 
that is commonly used in a CSR. The examples are in compliance of 
[FDA's Portable Document Format (PDF) Specifications](https://www.fda.gov/media/76797/download)

> FDA's PDF specification is a general reference, each organization may define 
> more specific TLF format requirement that can be different from the examples in this book.

These TLFs are typically used to summarize efficacy and safety statistical analysis results of a drug (or treatment).

## Table for Baseline Characteristics

```{r}
library(table1)
library(r2rtf)
data(r2rtf_adsl)

tbl <- table1(~ SEX + AGE + RACE | TRT01P, data = r2rtf_adsl)
tbl
```

```{r}
tbl %>% as.data.frame() %>% 
        rtf_body() %>% 
        rtf_encode() %>% 
        write_rtf("tlf/bs-1.rtf")
```

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
r2rtf:::rtf_convert_format("tlf/bs-1.rtf", output_dir = "tlf")
knitr::include_graphics("tlf/bs-1.pdf")
```
